Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2022-03-29 paper/tidy-missing-data-paper.Rmd
misc style fixes

**MCAR** is where values being missing have no association with observed or unobserved data, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}) = Pr(B = 1 | \psi)$. Essentially, the probability of an observation being missing is unrelated to anything else, only the parameter $\psi$, the overall probability of missingness. Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved. In **MAR**, missingness only depends on data observed, not data missing, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(B = 1 | Y_{obs}, \psi)$. Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness. In **MNAR**, missingness is related to values observed, and unobserved: $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi)$. This assumes conditioning on all observations: data goes missing due to some phenomena unobserved, including the structure of the missing data itself. This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena. Visualizations can help assess whether data may be MCAR, MAR or MNAR.

# Existing Software {#existing-software}

@@

### scan: searching for common missing value labels {#verbs-search}

@@

### replace: replacing values with missing values {#verbs-replace-with}

@@

### add: adding missingness summary variables {#verbs-add-cols}

@@

### shadow: creating nabular data {#verbs-nabular}

@@

### flag: describing different types of missing values {#verbs-recode}

@@

### impute: imputing values {#verbs-impute}

@@

### track: shadow and impute missing values {#verbs-track}

>>>>

MCAR is where values being missing have no association with observed or unobserved data, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}) = Pr(B = 1 | \psi)$. Essentially, the probability of an observation being missing is unrelated to anything else, only the parameter $\psi$, the overall probability of missingness. Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved. In MAR, missingness only depends on data observed, not data missing, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(B = 1 | Y_{obs}, \psi)$. Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness. In MNAR, missingness is related to values observed, and unobserved: $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi)$. This assumes conditioning on all observations: data goes missing due to some phenomena unobserved, including the structure of the missing data itself. This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena. Visualizations can help assess whether data may be MCAR, MAR or MNAR.

# Existing software {#existing-software}

@@

### Scan: searching for common missing value labels {#verbs-search}

@@

### Replace: replacing values with missing values {#verbs-replace-with}

@@

### Add: adding missingness summary variables {#verbs-add-cols}

@@

### Shadow: creating nabular data {#verbs-nabular}

@@

### Flag: describing different types of missing values {#verbs-recode}

@@

### Impute: imputing values {#verbs-impute}

@@

### Track: shadow and impute missing values {#verbs-track}
